ChemKED: a human- and machine-readable data standard for chemical kinetics experiments

### 2017-03-31 pyked.tex
edits through the end of the papre

@@

The combustion community has not yet widely adopted a standard format for
experimental data, but some researchers and groups have proposed solutions to
this problem. Most notably, Frenklach developed the PrIMe (\textbf{Pr}ocess
\textbf{I}nformatics \textbf{M}od\textbf{e}l) data
format~\autocite{Frenklach:2007bm,You:2011hy} and an associated online database,
the PrIMe Data Warehouse~\autocite{PrIMe}. PrIMe files encode fundamental
combustion experimental data, along with kinetic models and calculated
quantities, using the Extensible Markup Language (XML) standard.

@@

More recently, Varga et al.~\autocite{Varga2015a,Varga2015b} developed the
ReSpecTh standard, which builds on the PrIMe format. ReSpecTh adds important
features that make files better standalone representations of experimental
data---i.e., more informative by themselves rather than in concert with a larger
system. For example, where PrIMe uses internal bibliographic references,
ReSpecTh adds a field for typical bibliographic data. The ReSpecTh standard also
provides machine-readable formats for specifying ignition experiments, including
a field for the definition of autoignition and the ability to specify
facility-specific effects. However, ReSpecTh experimental files are another
XML-based format, and as such, suffer from the same usability issues as PrIMe.
Moreover, the standard remains closed, requiring potential users to register
with the standard's authors to access the specification.

@@

This section will provide a brief introduction to the \ck{} database format.
This is not a complete listing of available fields and more complete examples of
\ck{} files can be found in the \pybox{tests} directory of the \pk{} package. In
addition, in Section~\ref{sec:usage-example}, we will provide examples of \ck{}
files. Note that in the following, text with gray background is a \yabox{yaml}
keyword or value, while text with a yellow background is \pybox{Python} code.

\ck{} files use the YAML data serialization format. This format offers the
advantages of being human readable, written in plain text, and having parsers in
most common programming languages, including Python, C++, Java, Perl, MATLAB,
and many others. The YAML syntax is quite simple. The basic file structure is
made of mappings, delimited by a colon. The key for the mapping is on the left
of the colon, and the value on the right can be a single value, a sequence of
values, a nested mapping, or some combination of these:
%
\begin{yamlbox}
key1: value  # Single-value mapping
key2:  # Sequence format
  - value
  - value
key3:  # Nested mapping
  key4: 0
key5:  # Sequence of mappings
  - key-s1: 0
    key-s2: value
  - key-s2: value
    key-s1: 0
\end{yamlbox}

The \yabox{value} can be a string, integer, or floating point number. The \ck{}
format is designed to include all of the information necessary to simulate a
given experiment and can be broken into two main sections. The first section
details all of the ``meta'' information about the experiment and the \ck{} file.
In this section, the common \yabox{author} mapping is used for any fields where
an author must be specified. The \yabox{author} mapping contains the following
fields:
%
\begin{itemize}
    \item \yabox{name} (required, string): The author's full name
    \item \yabox{ORCID} (optional, string): the author's ORCID number
\end{itemize}
%
More generally for the meta information, the required keys in this section
are:
%
\begin{itemize}
    \item \yabox{file-author} (required, \yabox{author}-type mapping): The
    author of the \ck{} file
    \item \yabox{file-version} (required, integer): The version of the \ck{}
    file
    \item \yabox{chemked-version} (required, string): The version of the \ck{}
    schema that this file conforms to
    \item \yabox{reference} (required, mapping): The reference information for
    the article associated with the data in the file. The fields in this
    mapping are:
    \begin{itemize}
        \item \yabox{authors} (required, sequence): A sequence of \yabox{author}
        mappings
        \item \yabox{journal} (required, string): The name of the publishing
        journal
        \item \yabox{year} (required, integer): The year of publication
    \end{itemize}
    \item \yabox{experiment-type} (required, string): Currently, only
    \yabox{ignition delay} is supported
    \item \yabox{apparatus} (required, string): The type of apparatus used to
    perform the experiment
\end{itemize}

The second section of the file encodes the experimental data. In the \ck{} file,
this is done in a sequence of mappings, with the following information required
in each element of the sequence:
%
\begin{itemize}
    \item \yabox{temperature} (required, sequence): The temperature of the
    experiment, with units
    \item \yabox{ignition-delay} (required, sequence): The ignition delay of the
    experiment, with units
    \item \yabox{pressure} (required, string): The pressure of the experiment,
    with units
    \item \yabox{composition} (required, sequence): The composition of the
    mixture in the experiment, a sequence of mappings with the following fields
    \begin{itemize}
        \item \yabox{species-name} (required, string): The name of the species
        \item \yabox{mole-fraction}, \yabox{mass-fraction}, \yabox{mole-percent}
        (required, float): The mole fraction or percent, or the mass fraction.
        Only one of the three is permitted in a given \yabox{composition}
        sequence.
        \item \yabox{InChI}, \yabox{SMILES}, \yabox{atomic-composition}, or
        \yabox{elemental-composition} (required, string or sequence): The InChI
        or SMILES representing the molecule, or its atomic or elemental
        composition (\yabox{atomic-composition} and
        \yabox{elemental-composition} are synonyms) as a sequence
    \end{itemize}
    \item \yabox{ignition-type} (required, mapping); The type of ignition
    measured in the experiments. The required fields are
    \begin{itemize}
        \item \yabox{type} (required, string): How the ignition delay was
        measured, one of \yabox{d/dt max} to indicate the ignition point is
        found at the maximum of the time derivative of the \yabox{target},
        \yabox{max} or \yabox{min} to indicate the ignition point is at the
        maximum or minimum of the \yabox{target}, or \yabox{1/2 max} to
        indicate the half-maximum point
        \item \yabox{target} (required, string): The target for the type of the
        ignition point, one of \yabox{temperature}, \yabox{pressure},
        \yabox{OH}, \yabox{OH*}, \yabox{CH}, or \yabox{CH*}
    \end{itemize}
\end{itemize}

@@

The basic user interface to the \ck{} file in \pk{} is through the
\pybox{ChemKED} class. The \pybox{ChemKED} class constructor takes the file name
of a \ck{} file or a Python dictionary containing the contents of a \ck{} file
as its argument. When the file or dictionary is loaded, it is validated to have
the correct format by the Python package Cerberus \autocite{cerberus}. The
schema used for validation of the \ck{} files is available publicly in the
source code repository for
PyKED\footnote{\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/chemked_schema.yaml}}.

@@

The \pybox{DataPoint} class has two instance methods defined, called
\pybox{get_cantera_mole_fraction()} and \pybox{get_cantera_mass_fraction()}.
With these methods, the composition of the reactant mixture can be output to a
format that can be used to set the composition of a Cantera~\pybox{Solution}.
Note that the \pybox{composition} specification does not contain the molecular
weights of the components, so conversion between mole fractions and mass
fractions is not possible.

The \pybox{ChemKED} class has one instance method defined:
\pybox{get_dataframe()}. Using this method, the user can get an instance of a
Pandas~\autocite{pandas} \pybox{DataFrame} that contains the information in the
list of \pybox{DataPoint}s. The user can specify the columns included in the
\pybox{DataFrame} by passing a list of column names to the
\pybox{output_columns} argument of the \pybox{get_dataframe()} method. The
possible columns are not case-sensitive and are:

@@

\pk{}~\autocite{PyKED} relies on well-established scientific Python software tools.
These include NumPy~\autocite{vanderWalt:2011np} and Pandas~\autocite{pandas,McKinney2010}
for array manipulation, Pint~\autocite{Grecco2016} for interpreting and converting
between units, \texttt{ruamel.yaml}~\autocite{ruamel_yaml} for parsing YAML
files, Cerberus for validating \ck{} files~\autocite{cerberus}, and
pytest~\autocite{pytest:3.0.1} for unit testing. Travis-CI~\autocite{Travis2016}
provides continuous integration testing.

@@

The \ck{} file that will be used in this example can be found in the
\pybox{tests} directory of the PyKED
repository\footnote{\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/tests/testfile_rcm.yaml}}.
Examining that file, we find the first section specifies the information about
the \ck{} file itself:
%
\begin{yamlbox}
file-author:
  name: Kyle E Niemeyer
  ORCID: 0000-0003-4425-7097
file-version: (1, 0)
chemked-version: 0.0.1
\end{yamlbox}
%
Then, we find the information regarding the article in the literature from which
this data was taken. In this case, the data point is from the work of
\textcite{Mittal2006a}:
%
\begin{yamlbox}
reference:
  doi: 10.1002/kin.20180
  authors:
    - name: Gaurav Mittal
    - name: Chih-Jen Sung
      ORCID: 0000-0003-2046-8076
    - name: Richard A Yetter
  journal: International Journal of Chemical Kinetics
  year: 2006
  volume: 38
  pages: 516-529
  detail: Fig. 6, open circle
experiment-type: ignition delay
apparatus:
  kind: rapid compression machine
  institution: Case Western Reserve University
  facility: CWRU RCM
\end{yamlbox}
%
Finally, there is just a single datapoint in this file, which codes the ignition
delay for the experiment, the mixture composition, the initial temperature,
pressure, compression time, ignition type, and the volume history that specifies
how the volume of the reactor varies with time, simulating the compression
stroke and post-compression processes.
%
\begin{yamlbox}
datapoints:
  - temperature: 297.4 kelvin
    ignition-delay: 1.0 ms
    pressure: 958.0 torr
    composition:
      - species-name: H2
        InChI: 1S/H2/h1H
        mole-fraction: 0.12500
      - species-name: O2
        InChI: 1S/O2/c1-2
        mole-fraction: 0.06250
      - species-name: N2
        InChI: 1S/N2/c1-2
        mole-fraction: 0.18125
      - species-name: Ar
        InChI: 1S/Ar
        mole-fraction: 0.63125
    ignition-type:
      target: pressure
      type: d/dt max
    compression-time: 38.0 ms
    volume-history:
      time:
        units: s
        column: 0
      volume:
        units: cm3
        column: 1
      values:
        - [0.00E+000, 5.47669375000E+002]
        - [1.00E-003, 5.46608789894E+002]
        - [2.00E-003, 5.43427034574E+002]
        - [3.00E-003, 5.38124109043E+002]
        ...
\end{yamlbox}
%
The values for the \yabox{volume-history} are truncated to save space. A user
might like to load the information from this file to perform a simulation to
calculate the ignition delay using Cantera. All of the information required to
perform this simulation is present in the \ck{} file, with the exception of a
chemical kinetic mechanism for H$_2$/CO combustion.
%
\begin{pythonbox}
import cantera as ct
from pyked import ChemKED

# Load the ChemKED file and retrieve the first element of the
# datapoints list, which is an instance of the DataPoint class
ck = ChemKED('testfile_rcm.yaml')
dp = ck.datapoints[0]
\end{pythonbox}
%
The initial temperature, pressure, and mixture composition can be read from the
instance of the \pybox{DataPoint} class. \pk{} uses Pint \pybox{Quantities} to
store values with units, while Cantera expects a floating point value in SI
units as input. Therefore, we use the built-in capabilities of Pint to convert
the units from those specified in the \ck{} file to SI units:
%
\begin{pythonbox}
T_0 = dp.temperature.to('K').magnitude
P_0 = dp.pressure.to('Pa').magnitude
X_0 = dp.get_cantera_mole_fraction()

# Load the mechanism and set the initial state of the mixture
gas = ct.Solution('h2-co-mechanism.cti')
gas.TPX = T_0, P_0, X_0

# Create the reactor and the outside environment
reac = ct.IdealGasReactor(gas)
env = ct.Reservoir(ct.Solution('air.xml'))
\end{pythonbox}
%
To apply the effect of the volume trace to the \pybox{IdealGasReactor}, a
\pybox{Wall} is installed between the reactor and the environment and assigned
a velocity. Although we do not show the details here, a reference implementation
of a class that computes a wall velocity given the volume history of the reactor
is available in CanSen~\autocite{cansen}, in the
\pybox{cansen.profile.VolumeProfile} class.
%
\begin{pythonbox}
time = dp.volume_history.time
volume = dp.volume_history.volume
ct.Wall(reac, env, velocity=VolumeProfile(time=time, volume=volume))

@@

The \ck{} file used in this example can be found in the \pybox{tests} directory of the \pk{}
repository\footnote{\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/tests/testfile_st_p5.yaml}}.
The data in this file comes from Stranic et al.~\autocite{Stranic:2012}, describing
shock-tube ignition delays for \textit{tert}-butanol. We have omitted the file
meta information below for space; the format is largely similar to the example in Section~\ref{sec:rcm-modeling}.
%
In this \ck{} file, multiple data points are specified with some common
conditions, including a common mixture composition and common definition of
ignition delay. Therefore, a \yabox{common-properties} section is specified.
%
\begin{yamlbox}
common-properties:
  composition: &comp
    - species-name: t-butanol
      InChI: 1S/C4H10O/c1-4(2,3)5/h5H,1-3H3
      mole-fraction: 0.003333333
    - species-name: O2
      InChI:  1S/O2/c1-2
      mole-fraction: 0.04
    - species-name: Ar
      InChI:  1S/Ar
      mole-fraction: 0.956666667
  ignition-type:  &ign
    target: OH* concentration
    type: 1/2 max
\end{yamlbox}
%
This block uses the ability of YAML files to define an anchor with the
\yabox{&} symbol and refer to that section later with the \yabox{*} symbol, as
can be seen in the definition of the \yabox{datapoints} section:
%
\begin{yamlbox}
datapoints:
  - temperature: 1459 kelvin
    ignition-delay: 347 us
    pressure: 1.60 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1389 kelvin
    ignition-delay: 756 us
    pressure: 1.67 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1497 kelvin
    ignition-delay: 212 us
    pressure: 1.55 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1562 kelvin
    ignition-delay: 105 us
    pressure: 1.50 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
\end{yamlbox}

In this example, we would like to run constant volume simulations at each of the
pressure and temperature conditions in the \yabox{datapoints} list. Once again,
the \ck{} file specifies all the information required for the simulations except
for the chemical kinetic mechanism.
%
\begin{pythonbox}
import cantera as ct
from pyked import ChemKED

@@

# Since the composition is specified by the common-properties,
# just take the composition of the first DataPoint
X_0 = ck.datapoints[0].get_cantera_mole_fraction()

# Loop through each of the DataPoints
for dp in ck.datapoints:
    T_0 = dp.temperature.to('K').magnitude
    P_0 = dp.pressure.to('Pa').magnitude
    gas.TPX = T_0, P_0, X_0
    reac = ct.IdealGasReactor(gas)
    netw = ct.ReactorNet([reac])
    # Define ignition delay as T_0 + 400 K
    while reac.T < T_0 + 400:
        netw.step()

    print('The ignition delay for T_0={}, P_0={} is: {} seconds'.format(
        T_0, P_0, netw.time
    ))
\end{pythonbox}
%
Note that the ignition delay in this example is defined as equal to $T_0 +
400\,\mathrm{K}$, a simplified definition used for this example. In general, the
user could post-process the concentration information from the simulation to
determine one-half the maximum OH* concentration to match the experimental
definition of ignition delay.
%====================================================================
\section{Conclusions and future work}

>>>>

\usepackage[version=4]{mhchem}

\usepackage{siunitx}
\sisetup{group-separator={,},
     detect-all,
     binary-units,
     list-units = single,
     range-units = single,
     tophrase = --,
     per-mode = symbol-or-fraction,
     separate-uncertainty = true,
     list-final-separator = {, and }
}
\DeclareSIUnit\atm{atm}

@@

The combustion community has not yet widely adopted a standard format for
experimental data, but some researchers and groups have proposed solutions to
this problem. Most notably, Frenklach developed the PrIMe (\textbf{Pr}ocess
\textbf{I}nformatics \textbf{M}od\textbf{e}l) data
format~\autocite{Frenklach:2007bm,You:2011hy} and an associated online database,
the PrIMe Data Warehouse.\footnote{\url{http://primekinetics.org}}
PrIMe files encode fundamental
combustion experimental data, along with kinetic models and calculated
quantities, using the Extensible Markup Language (XML) standard.

@@

More recently, Varga et al.~\autocite{Varga2015a,Varga2015b} developed the
ReSpecTh standard, which builds on the PrIMe format. ReSpecTh adds important
features that make files better standalone representations of experimental
data---i.e., more informative by themselves rather than in concert with a larger
system. For example, where PrIMe uses internal bibliographic references,
ReSpecTh adds a field for typical bibliographic data. The ReSpecTh standard also
provides machine-readable formats for specifying ignition experiments, including
a field for the definition of autoignition and the ability to specify
facility-specific effects. However, ReSpecTh experimental files are another
XML-based format, and, as such, suffer from the same usability issues as PrIMe.
Moreover, the standard remains closed, requiring potential users to register
with the standard's authors to access the specification.

@@

This section will provide a brief introduction to the \ck{} database format.
This is not a complete listing of available fields and more complete examples of
\ck{} files can be found in the \pybox{tests} directory of the \pk{} package. In
addition, in Section~\ref{sec:usage-example}, we will provide examples of \ck{}
files. We indicate \yabox{yaml} keywords or values using text with gray background,
and \pybox{Python} code using text with a yellow background.

\ck{} files use the YAML data serialization format~\autocite{yaml:1.2}. This
format offers the advantages of being human readable, written in plain text, and
having parsers in most common programming languages, including Python, C++, Java,
Perl, MATLAB, and many others. The YAML syntax is quite simple. The basic file
structure is made of mappings, delimited by a colon. The key for the mapping is
on the left of the colon, and the value on the right can be a single value,
a sequence of values, a nested mapping, or some combination of these:
%
\begin{yamlbox}
key1: value  # Single-value mapping
key2:  # Sequence format
  - value
  - value
key3:  # Nested mapping
  key4: 0
key5:  # Sequence of mappings
  - key-s1: 0
    key-s2: value
  - key-s2: value
    key-s1: 0
\end{yamlbox}

The \yabox{value} can be a string, integer, or floating point number. The \ck{}
format is designed to include all of the information necessary to simulate a
given experiment and can be broken into two main sections. The first section
details all of the ``meta'' information about the experiment and the \ck{} file.
In this section, the common \yabox{author} mapping is used for any fields where
an author must be specified. The \yabox{author} mapping contains the following
fields:
%
\begin{itemize}
    \item \yabox{name} (required, string): The author's full name
    \item \yabox{ORCID} (optional, string): the author's ORCID number
\end{itemize}
%
More generally for the meta information, the required keys in this section
are:
%
\begin{itemize}
    \item \yabox{file-author} (required, \yabox{author}-type mapping): The
    author of the \ck{} file
    \item \yabox{file-version} (required, integer): The version of the \ck{}
    file
    \item \yabox{chemked-version} (required, string): The version of the \ck{}
    schema that this file conforms to
    \item \yabox{reference} (required, mapping): The reference information for
    the article associated with the data in the file. The fields in this
    mapping are:
    \begin{itemize}
        \item \yabox{authors} (required, sequence): A sequence of \yabox{author}
        mappings
        \item \yabox{journal} (required, string): The name of the publishing
        journal
        \item \yabox{year} (required, integer): The year of publication
        \item \yabox{doi} (optional, string): The article DOI
        \item \yabox{volume} (required, integer): The journal volume number
        \item \yabox{pages} (optional, string): The article pages
        \item \yabox{detail} (optional, string): A description of from where the
        data originated (e.g., figure or table number)
    \end{itemize}
    \item \yabox{experiment-type} (required, string): Currently, only
    \yabox{ignition delay} is supported
    \item \yabox{apparatus} (required, string): The type of apparatus used to
    perform the experiment (\yabox{shock tube} or \yabox{rapid compression machine})
\end{itemize}

The second section of the file encodes the experimental data. In the \ck{} file,
this is done in a sequence of mappings, with the following information required
in each element of the sequence:
%
\begin{itemize}
    \item \yabox{temperature} (required, sequence): The temperature of the
    experiment, with units
    \item \yabox{ignition-delay} (required, sequence): The ignition delay of the
    experiment, with units
    \item \yabox{pressure} (required, string): The pressure of the experiment,
    with units
    \item \yabox{composition} (required, sequence): The composition of the
    mixture in the experiment, a sequence of mappings with the following fields
    \begin{itemize}
        \item \yabox{species-name} (required, string): The name of the species
        \item \yabox{mole-fraction}, \yabox{mass-fraction}, \yabox{mole-percent}
        (required, float): The mole fraction or percent, or the mass fraction.
        Only one of the three is permitted in a given \yabox{composition}
        sequence.
        \item \yabox{InChI}, \yabox{SMILES}, \yabox{atomic-composition}, or
        \yabox{elemental-composition} (required, string or sequence): The InChI
        or SMILES representing the molecule, or its atomic or elemental
        composition (\yabox{atomic-composition} and
        \yabox{elemental-composition} are synonyms) as a sequence
    \end{itemize}
    \item \yabox{ignition-type} (required, mapping): The type of ignition
    measured in the experiments. The required fields are
    \begin{itemize}
        \item \yabox{type} (required, string): How the ignition delay was
        measured; one of \yabox{d/dt max} to indicate the ignition point is
        found at the maximum of the time derivative of the \yabox{target},
        \yabox{max} or \yabox{min} to indicate the ignition point is at the
        maximum or minimum of the \yabox{target}, or \yabox{1/2 max} to
        indicate the half-maximum point
        \item \yabox{target} (required, string): The target for the type of the
        ignition point; one of \yabox{temperature}, \yabox{pressure},
        \yabox{OH}, \yabox{OH*}, \yabox{CH}, or \yabox{CH*}
    \end{itemize}
\end{itemize}

@@

The basic user interface to the \ck{} file in \pk{} is through the
\pybox{ChemKED} class. The \pybox{ChemKED} class constructor takes the filename
of a \ck{} file or a Python dictionary containing the contents of a \ck{} file
as its argument. When the file or dictionary is loaded, it is validated to have
the correct format by the Python package Cerberus \autocite{cerberus}. The
schema used for validation of the \ck{} files is available publicly in the
source code repository for
PyKED\footnote{\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/chemked_schema.yaml}}.

@@

The \pybox{DataPoint} class defines two instance methods:
\pybox{get_cantera_mole_fraction()} and \pybox{get_cantera_mass_fraction()}.
With these methods, the composition of the reactant mixture can be output to a
format that can be used to set the composition of a
Cantera~\pybox{Solution}~\autocite{Cantera:2.3.0}.
The \pybox{composition} specification does not contain the molecular
weights of the components, so conversion between mole fractions and mass
fractions is not currently possible.

The \pybox{ChemKED} class defines one instance method:
\pybox{get_dataframe()}. Using this method, the user can get an instance of a
Pandas \pybox{DataFrame}~\autocite{pandas} that contains the information in the
list of \pybox{DataPoint}s. The user can specify the columns included in the
\pybox{DataFrame} by passing a list of column names to the
\pybox{output_columns} argument of the \pybox{get_dataframe()} method. The
possible columns are not case-sensitive and are:

@@

\pk{}~\autocite{PyKED} relies on well-established scientific Python software tools.
These include NumPy~\autocite{vanderWalt:2011np} and Pandas~\autocite{pandas,McKinney2010}
for array manipulation, Pint~\autocite{Grecco2016} for interpreting and converting
between units, \texttt{ruamel.yaml}~\autocite{ruamel_yaml} for parsing YAML
files, Cerberus for validating \ck{} files~\autocite{cerberus}, and
pytest~\autocite{pytest:3.0.1} for unit testing.
Travis-CI\footnote{\url{https://travis-ci.org/}} provides continuous integration
testing.

@@

The \ck{} file that will be used in this example can be found in the
\pybox{tests} directory of the PyKED
repository\footnote{\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/tests/testfile_rcm.yaml}}.
Examining that file, we find the first section specifies the information about
the \ck{} file itself:
%
\begin{yamlbox}
file-author:
  name: Kyle E Niemeyer
  ORCID: 0000-0003-4425-7097
file-version: (1, 0)
chemked-version: 0.0.1
\end{yamlbox}
%
Then, we find the information regarding the article in the literature from which
this data was taken. In this case, the dataset comes from the work of
\textcite{Mittal2006a}:
%
\begin{yamlbox}
reference:
  doi: 10.1002/kin.20180
  authors:
    - name: Gaurav Mittal
    - name: Chih-Jen Sung
      ORCID: 0000-0003-2046-8076
    - name: Richard A Yetter
  journal: International Journal of Chemical Kinetics
  year: 2006
  volume: 38
  pages: 516-529
  detail: Fig. 6, open circle
experiment-type: ignition delay
apparatus:
  kind: rapid compression machine
  institution: Case Western Reserve University
  facility: CWRU RCM
\end{yamlbox}
%
Finally, there is just a single datapoint in this file, which describes the ignition
delay for the experiment, the mixture composition, the initial temperature,
pressure, compression time, ignition type, and the volume history that specifies
how the volume of the reactor varies with time, for simulating the compression
stroke and post-compression processes.
%
\begin{yamlbox}
datapoints:
  - temperature: 297.4 kelvin
    ignition-delay: 1.0 ms
    pressure: 958.0 torr
    composition:
      - species-name: H2
        InChI: 1S/H2/h1H
        mole-fraction: 0.12500
      - species-name: O2
        InChI: 1S/O2/c1-2
        mole-fraction: 0.06250
      - species-name: N2
        InChI: 1S/N2/c1-2
        mole-fraction: 0.18125
      - species-name: Ar
        InChI: 1S/Ar
        mole-fraction: 0.63125
    ignition-type:
      target: pressure
      type: d/dt max
    compression-time: 38.0 ms
    volume-history:
      time:
        units: s
        column: 0
      volume:
        units: cm3
        column: 1
      values:
        - [0.00E+000, 5.47669375000E+002]
        - [1.00E-003, 5.46608789894E+002]
        - [2.00E-003, 5.43427034574E+002]
        - [3.00E-003, 5.38124109043E+002]
        ...
\end{yamlbox}
%
The values for the \yabox{volume-history} are truncated here to save space. A user
might want to load the information from this file to perform a simulation using
Cantera to calculate the ignition delay. All of the information required to
perform this simulation is present in the \ck{} file, with the exception of a
chemical kinetic model for \ce{H2}\slash {CO} combustion.
%
\begin{pythonbox}
import cantera as ct
from pyked import ChemKED

# Load the ChemKED file and retrieve the first element of the
# datapoints list, which is an instance of the DataPoint class
ck = ChemKED('testfile_rcm.yaml')
dp = ck.datapoints[0]
\end{pythonbox}
%
The initial temperature, pressure, and mixture composition can be read from the
instance of the \pybox{DataPoint} class. \pk{} uses Pint \pybox{Quantities} to
store values with units, while Cantera expects a floating point value in SI
units as input. Therefore, we use the built-in capabilities of Pint to convert
the units from those specified in the \ck{} file to SI units:
%
\begin{pythonbox}
T_initial = dp.temperature.to('K').magnitude
P_initial = dp.pressure.to('Pa').magnitude
X_initial = dp.get_cantera_mole_fraction()

# Load the mechanism and set the initial state of the mixture
gas = ct.Solution('h2-co-mechanism.cti')
gas.TPX = T_initial, P_initial, X_initial

# Create the reactor and the outside environment
reac = ct.IdealGasReactor(gas)
env = ct.Reservoir(ct.Solution('air.xml'))
\end{pythonbox}
%
To apply the effect of the volume trace to the \pybox{IdealGasReactor}, a
\pybox{Wall} is installed between the reactor and the environment and assigned
a velocity. Although we do not show the details here, a reference implementation
of a class that computes a wall velocity given the volume history of the reactor
is available in CanSen~\autocite{cansen}, in the
\pybox{cansen.profile.VolumeProfile} class. Then, a Cantera \pybox{ReactorNet}
can be used to advance the state through autoignition, in this case to an end
time of \SI{50}{\milli\second}:
%
\begin{pythonbox}
time = dp.volume_history.time
volume = dp.volume_history.volume
ct.Wall(reac, env, velocity=VolumeProfile(time=time, volume=volume))

@@

The \ck{} file used in this example can be found in the \pybox{tests} directory of the \pk{}
repository\footnote{\url{https://github.com/pr-omethe-us/PyKED/blob/master/pyked/tests/testfile_st_p5.yaml}}.
The data in this file comes from Stranic et al.~\autocite{Stranic:2012}, describing
shock-tube ignition delays for \textit{tert}-butanol. We have omitted the file
meta information below for space; the format is largely similar to the example in Section~\ref{sec:rcm-modeling}.
%
This \ck{} file specifies multiple data points with some common
conditions, including a common mixture composition and common definition of
ignition delay. Therefore, a \yabox{common-properties} section is specified.
%
\begin{yamlbox}
common-properties:
  composition: &comp
    - species-name: t-butanol
      InChI: 1S/C4H10O/c1-4(2,3)5/h5H,1-3H3
      mole-fraction: 0.003333333
    - species-name: O2
      InChI:  1S/O2/c1-2
      mole-fraction: 0.04
    - species-name: Ar
      InChI:  1S/Ar
      mole-fraction: 0.956666667
  ignition-type:  &ign
    target: OH* concentration
    type: 1/2 max
\end{yamlbox}
%
This block uses the ability of YAML files to define an anchor with the
\yabox{&} symbol and refer to that section later with the \yabox{*} symbol, as
can be seen in the definition of the \yabox{datapoints} section:
%
\begin{yamlbox}
datapoints:
  - temperature: 1459 kelvin
    ignition-delay: 347 us
    pressure: 1.60 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1389 kelvin
    ignition-delay: 756 us
    pressure: 1.67 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1497 kelvin
    ignition-delay: 212 us
    pressure: 1.55 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
  - temperature: 1562 kelvin
    ignition-delay: 105 us
    pressure: 1.50 atm
    composition: *comp
    ignition-type: *ign
    equivalence-ratio: 0.5
\end{yamlbox}
The \yabox{common-properties} section is not required, but can save space and
help avoid errors when all data points share some common values.

In this example, we would like to run constant-volume simulations at each of the
pressure and temperature conditions in the \yabox{datapoints} list. Once again,
the \ck{} file specifies all the information required for the simulations except
for the chemical kinetic model, and Cantera can be used to simulate autoignition:
%
\begin{pythonbox}
import cantera as ct
from pyked import ChemKED

@@

# Since the composition is specified by the common-properties,
# just take the composition of the first DataPoint
X_initial = ck.datapoints[0].get_cantera_mole_fraction()

# Loop through each of the DataPoints
for dp in ck.datapoints:
    T_initial = dp.temperature.to('K').magnitude
    P_initial = dp.pressure.to('Pa').magnitude
    gas.TPX = T_initial, P_initial, X_initial
    reac = ct.IdealGasReactor(gas)
    netw = ct.ReactorNet([reac])
    # Define ignition delay as T_initial + 400 K
    while reac.T < T_initial + 400:
        netw.step()

    print('The ignition delay for T_initial={}, P_initial={} is: '
          '{} seconds'.format(T_initial, P_initial, netw.time)
          )
\end{pythonbox}
%
The ignition delay in this example is defined as equal to the \SI{400}{\kelvin}
above the initial temperature, a simplified definition used for this example.
In general, the user could post-process the concentration information from the
simulation to determine one-half the maximum \ce{OH^*} concentration to match
the experimental definition of ignition delay.
%====================================================================
\section{Conclusions and future work}
